Evaluative adjective sentences: \\ A question-based analysis of projection

### 2019-04-22 paper/paper-anon.tex
updated bibliography file
jd added paper ocmments

\newcommand{\jt}[1]{\textbf{\color{blue}JT: #1}}
\newcommand{\jd}[1]{\textbf{\color{green}[jd: #1]}}

@@

Two contents of evaluative adjective sentences, like {\em Kim was smart to watch the movie}, are the prejacent, that Kim watched the movie, and the generalization, that the degree to which Kim watching the movie is smart is higher than the contextual standard of {\em smart}. The prejacent is standardly analyzed as a presupposition
(e.g., \citealt{norrick78,barker02,oshima09b,kertz2010}). This paper argues against such analyses because, among other things, they do not capture an interaction between the prejacent and the generalization that has not yet been observed for projective content: when the prejacent projects, the generalization does not, and when the prejacent does not project, the generalization does. We develop an analysis according to which the prejacent is not a lexically specified presupposition but projects if it is not at-issue with respect to the question addressed by the utterance of the evaluative adjective sentence. In addition to capturing the interaction between the prejacent and the generalization, our question-based projection analysis significantly expands on previous such analyses (e.g., \citealt{beaver-clark08,best-question,brst-ar,abrusan2011}) by incorporating a novel constraint on the question addressed by an utterance: the more the truth of an utterance content follows from the common ground, the less likely the question is about that content. We provide experimental evidence for the proposed analysis and argue that our analysis improves on that of \citealt{karttunen-etal2014}, according to which evaluative adjectives are systematically ambiguous.

@@

We discuss below why such examples are problematic for analyses of the prejacent as a lexically specified presupposition. But we first need to mention that there is variation in the population of native speakers of American English: whereas all speakers are able to interpret EASs in which the prejacent does not project, i.e., are able to retrieve the intended interpretations of examples like (\ref{nat}), a sizeable portion strongly prefers to realize such interpretations with variants that includes {\em enough}, as in {\em I wasn't stupid enough to go stumbling through the junkyard in the dark} for (\ref{nat}b); see also \citealt{karttunen2013} and \citealt{karttunen-etal2014} for this observation. Nevertheless, EASs in which the prejacent does not project are part of American English, as evidenced by the existence of naturally occurring examples like (\ref{nat}); further  such examples are presented in this paper. Additional evidence comes from the fact that there are native speakers who judge such EASs to be perfectly acceptable\footnote{The audiences to which this research was presented over the years always included native speakers that judged examples like (\ref{nat}) to be acceptable. More systematic evidence for the existence of such speakers comes from the acceptability rating study presented in Appendix \ref{s-acc}: of the 94 self-reported native speakers of American English that participated in the study, about 20-30\% judged negated evaluative adjective sentences without {\em enough} to be acceptable under a non-projecting interpretation of the prejacent.} and who produce examples like (\ref{nat}). We therefore assume that all native speakers of American English are able to retrieve both interpretations of EASs even if they might not produce EASs in which the prejacent does not project. Our goal in this paper is to analyze the interpretation of EASs; we briefly return to the observed variation after developing our analysis.

@@

As mentioned above, previous analyses of EAS have attributed the projectivity of the prejacent of EAS to its lexical specification as a presupposition. More recently, the observation that some utterance content is not highly projective and sensitive to the discourse context has led to the development of analyses according to which the projectivity of content is not analyzed as derived from a lexical specification but rather from being backgrounded or not-at-issue content (e.g., \citealt{abrusan2011,abrusan2016,brst-salt10,best-question,brst-ar}). In this paper, we follow such proposals and argue that the prejacent of an EAS projects when it is not at-issue with respect to the question addressed by the utterance of the EAS. Empirical motivation for developing such an analysis comes from the observation that the prejacent of EAS is not highly projective.  AUTHORS' 2018 experimental investigation found that the projectivity of the prejacent of EASs with {\em stupid} was significantly lower than that of non-restrictive relative clauses and appositive content, as well as of the content of the complement of the factive predicates {\em be annoyed} and {\em know}. In naturally occurring data, too, the prejacent of NEASs is not highly projective. In a corpus-based web study, we collected `certainty' ratings on a 7-point Likert scale from 226 native speakers of American English for the prejacents of 59 naturally occurring NEAS: the higher the certainty rating, the more projective the prejacent. Figure \ref{f-corpus} shows the mean certainty ratings of the 59 NEASs by evaluative adjective: across the 59 examples, the prejacent is not highly projective; the mean certainty rating was only 3.2. (Details on this study are given in Appendix \ref{a-corpus}.)\footnote{Although this paper  largely considers EAS embedded under negation, the examples in (i) show that when an EAS is embedded under other entailment-canceling operators, such as the antecedent of a conditional or the epistemic possibility modal {\em perhaps}, the prejacent also need not project.

\begin{exe}
\exi{(i)}
\begin{xlist}
\ex 
\begin{xlist}
\exi{A:} I wonder how many people would be banned for using SAM on Killing Floor.
\exi{B:} If they were stupid to leave it running then maybe a few.\\ \url{steamcommunity.com/app/730/discussions/0/540744934462316309/}
\end{xlist}
\ex I am seaching [sic] for a remote cabine in Finland that is available for renting. [...]
Region wise I would prefer Lapland or Lakeside. Perhaps we are lucky to see northern lights. \\ \url{www.tripadvisor.com/ShowTopic-g189896-i442-k11448709-Remote rental cottage-Finland.html}

\end{xlist}
\end{exe}
}

@@

\end{exe}} For empirical evidence that at-issueness predicts projection, see AUTHORS 2018, who investigated the hypothesis for 19 projective contents associated with American English expressions.

@@

The proposal not only accounts for the projectivity of the prejacent of EAS but also for the interaction in the projection of the prejacent and the generalization. Given that both the prejacent and the generalization can be at-issue, but exactly one of them may be at-issue in any given utterance of an EAS, it falls out from the analysis that if the prejacent projects, the generalization does not, and vice versa: if the prejacent is not at-issue, it projects, and hence the generalization must be at-issue and doesn't project.

@@

The next section provides experimental evidence in support of the question-based projection analyses of EAS. Specifically, it presents the findings of two experiments designed to investigate two predictions of the analysis: the first prediction is that the stronger the inference to the truth of the generalization from the common ground, the less likely the prejacent is taken to project. The second prediction is that the stronger the inference to the truth of the generalization from the common ground, the more likely the prejacent is taken to be not at-issue.

@@

The question-based projection analysis we developed in the previous section leads us to expect that the more the prejacent or the generalization are not at-issue, the more they project. Preliminary evidence that the at-issueness of the prejacent influences its projection comes from AUTHORS 2018. One of the contents they investigated was the prejacent of EASs with {\em stupid} embedded under the polar question operator. The four evaluative adjective polar questions studied in their work are given in (\ref{stupid2}):

@@

Preliminary evidence for this prediction comes from an experiment reported on in \citealt{karttunen-etal2014}. In that experiment, participants were presented with written NEASs like those in (\ref{paul}), referred to as `statements', and they were asked whether the author of the statement believes the prejacent or the negation of the prejacent; a third response option was `cannot decide'. The experiment included one triple like that in (\ref{paul}) for each of the 19 evaluative adjectives tested ({\em arrogant, brave, careless, cruel, evil,
foolish, fortunate, heroic, humble, lucky, mean, nice, polite, rude,
sensible, smart, stupid, sweet, wise}). Each triple included a NEAS referred to by \citet{karttunen-etal2014} as `consonant', which means that ``there is a predisposition to assume or grant that for NP to VP would be Adj'' (p.237). For instance, the NEAS in (\ref{paul}a) is consonant because for Paul to take the best piece is smart; this is comparable to what we have characterized as the generalization following from the common ground. Each triple also included a NEAS that \citet{karttunen-etal2014} referred to as `dissonant', which means that ``there is a predisposition to assume or grant that for NP to VP would not be Adj'' ({\em ibid}). The NEAS in (\ref{paul}c) is dissonant because for the man to take the worst piece is not smart; this is comparable to what we have characterized as the falsity of the generalization following from the common ground. Finally, the third NEAS in each triple was considered `neutral', i.e., neither consonant or dissonant, like (\ref{paul}b).

@@

A second goal of Exp.~1 was to explore the prediction of our analysis that the discourse status of the generalization is influenced not just by the truth-conditional content of the EAS but also by contextual information. For this reason, Exp.~1 manipulated whether the critical information about the discourse status of the generalization came from the EAS or from context. Exp.~1 collected participants' ratings of the projection of the prejacents of NEASs. Specifically, the experiment collected gradient projection ratings, under the assumption that projection is a gradient property of utterance content, because speakers can be taken to be more or less committed to the truth of a particular content (AUTHORS 2018).\footnote{As discussed in AUTHORS: 498f., there are at least two interpretations of what it means for projectivity to be a gradient property: ``On a first interpretation, a listener's (or reader's) judgment that a content is projective to a certain extent means that the listener takes the speaker (or writer) to be committed to the content to that extent. On this interpretation, projectivity being a gradient property is a consequence of speaker commitment being a gradient property. On a second interpretation, a listener's judgment that a content is projective to a certain extent reflects the probability with which they believe the speaker to be committed to the content. On this interpretation, speaker commitment may be a binary, categorical property and projection variability arises from the listener?s uncertainty about the whether the speaker is committed." Like AUTHORS 2018, we remain agnostic about the underlying interpretation of projectivity as a gradient property, though our discussion of projection variability will be in line with the first interpretation.}

@@

To assess whether participants were attending to the task, the same 6 control stimuli were added to each list, for a total of 16 stimuli per list. As shown in (\ref{control-exp1}), the control stimuli also consisted of two-sentence discourses: the content of interest in (\ref{control-exp1}a-c) was hypothesized to be true, and the content of interest in (\ref{control-exp1}d-f) was hypothesized to be false.

@@

According to the analysis of EASs developed in section \ref{s2}, the prejacent is not a lexically-specified presupposition but projects when it is not at-issue with respect to the Discourse Question of the utterance. The analysis predicts that the more the truth of the generalization follows from the common ground, the more likely it is that the Discourse Question is not about the generalization, but about the prejacent, and, hence, the more likely the prejacent is to not project. In this section, we provided empirical evidence for two predictions of this analysis:  Exp.~1 showed that the prejacent is more likely to project when the truth of the generalization does not follow from the common ground than when the truth of the generalization does not follow from the common ground, and Exp.~2 showed that the prejacent is more likely to be at-issue when the truth of the generalization follows from the common ground than when the truth of the generalization does not follow from the common ground.

@@

\caption{Count of participants in Experiment 1 by their mean certainty ratings for the two types of NEASs}\label{f-dialect}
\end{center}
\end{figure}

@@

\bibliographystyle{cslipubs-natbib}
\bibliography{/Users/tonhauser.1/Documents/bibliography}

>>>>

\newcommand{\jt}[1]{\textbf{\color{blue}JT: #1}}
\newcommand{\jd}[1]{\textbf{\color{red}[jd: #1]}}

@@

Two contents of evaluative adjective sentences, like {\em Kim was smart to watch the movie}, are the prejacent (that Kim watched the movie) and the generalization (that the degree to which Kim watching the movie is smart is higher than the contextual standard of {\em smart}). The prejacent is standardly analyzed as a presupposition
(e.g., \citealt{norrick78,barker02,oshima09b,kertz2010}). This paper argues against such \jd{such = what?} analyses because, among other things, they do not capture an interaction between the prejacent and the generalization that has not yet been observed for projective content: when the prejacent projects, the generalization does not, and when the prejacent does not project, the generalization does \jd{does the evidence really favor such categorical a view?}. We develop an analysis according to which the prejacent is not a lexically specified presupposition but projects if it is not at-issue with respect to the question addressed by the utterance of the evaluative adjective sentence . In addition to capturing the interaction between the prejacent and the generalization, our question-based projection analysis significantly expands on previous such analyses (e.g., \citealt{beaver-clark08,best-question,brst-ar,abrusan2011}) by incorporating a novel constraint on the question addressed by an utterance: the more the truth of an utterance content follows from the common ground \jd{can something follow more or less, or should this instead be "the more likely the truth of an utterance content is to follow from the common ground..."}, the less likely the question is about that content. We provide experimental evidence for the proposed analysis and argue that our analysis improves on that of \citealt{karttunen-etal2014}, according to which evaluative adjectives are systematically ambiguous.

@@

We discuss below why such examples are problematic for analyses of the prejacent as a lexically specified presupposition. But we first need to mention \jd{"mention" sounds very offhand here. perhaps instead "Before discussing why such examples are problematic for..., we first would like to acknowledge that there is variation..."} that there is variation in the population of native speakers of American English: whereas all speakers are able to interpret EASs in which the prejacent does not project, i.e., are able to retrieve the intended interpretations of examples like (\ref{nat}), a sizeable portion strongly prefers to realize such interpretations with variants that includes {\em enough}, as in {\em I wasn't stupid enough to go stumbling through the junkyard in the dark} for (\ref{nat}b); see also \citealt{karttunen2013} and \citealt{karttunen-etal2014} for this observation. Nevertheless, EASs in which the prejacent does not project are part of American English, as evidenced by the existence of naturally occurring examples like (\ref{nat}); further  such examples are presented in this paper. Additional evidence comes from the fact that there are native speakers who judge such EASs to be perfectly acceptable\footnote{The audiences to which this research was presented over the years always included native speakers that judged examples like (\ref{nat}) to be acceptable. More systematic evidence for the existence of such speakers comes from the acceptability rating study presented in Appendix \ref{s-acc}: of the 94 self-reported native speakers of American English that participated in the study, about 20-30\% judged negated evaluative adjective sentences without {\em enough} to be acceptable under a non-projecting interpretation of the prejacent.} and who produce examples like (\ref{nat}). We therefore assume that all \jd{really "all"? is it necessary to assume that? it might be wrong} native speakers of American English are able to retrieve both interpretations of EASs even if they might not produce EASs in which the prejacent does not project. Our goal in this paper is to analyze the interpretation of EASs; we briefly return to the observed variation after developing our analysis.

@@

As mentioned above, previous analyses of EAS have attributed the projectivity of the prejacent of EAS to its lexical specification as a presupposition. More recently, the observation that some utterance content is not highly projective and sensitive to the discourse context has led to the development of analyses according to which the projectivity of content is not analyzed as derived from a lexical specification but rather from being backgrounded or not-at-issue content (e.g., \citealt{abrusan2011,abrusan2016,brst-salt10,best-question,brst-ar}). In this paper, we follow such proposals and argue that the prejacent of an EAS projects when it is not at-issue with respect to the question addressed by the utterance of the EAS. Empirical motivation for developing such an analysis comes from the observation that the prejacent of EAS is not highly projective.  AUTHORS' \jd{can just cite and not say "we", no?} 2018 experimental investigation found that the projectivity of the prejacent of EASs with {\em stupid} was significantly lower than that of non-restrictive relative clauses and appositive content, as well as of the content of the complement of the factive predicates {\em be annoyed} and {\em know}. In naturally occurring data, too, the prejacent of NEASs is not highly projective. In a corpus-based web study, we collected `certainty' ratings on a 7-point Likert scale from 226 native speakers of American English for the prejacents of 59 naturally occurring NEAS: the higher the certainty rating, the more projective the prejacent. Figure \ref{f-corpus} shows the mean certainty ratings of the 59 NEASs by evaluative adjective: across the 59 examples, the prejacent is not highly projective; the mean certainty rating was only 3.2. Details on this study are provided in Appendix \ref{a-corpus}.\footnote{Although this paper  largely considers EAS embedded under negation, the examples in (i) show that when an EAS is embedded under other entailment-canceling operators, such as the antecedent of a conditional or the epistemic possibility modal {\em perhaps}, the prejacent also need not project.

\begin{exe}
\exi{(i)}
\begin{xlist}
\ex 
\begin{xlist}
\exi{A:} I wonder how many people would be banned for using SAM on Killing Floor.
\exi{B:} If they were stupid to leave it running then maybe a few.\\ \url{steamcommunity.com/app/730/discussions/0/540744934462316309/}
\end{xlist}
\ex I am seaching [sic] for a remote cabine in Finland that is available for renting. [...]
Region wise I would prefer Lapland or Lakeside. Perhaps we are lucky to see northern lights. \\ \url{www.tripadvisor.com/ShowTopic-g189896-i442-k11448709-Remote rental cottage-Finland.html}
\end{xlist}
\end{exe}
 \jd{hm, example b) sounds very L2 overall (the nature of the typos, "cabine"; and the "we are" instead of "we'll be"). and that it's from tripadvisor further increases the prob that it's not from a native english speaker, which would be a shot in the foot i think}
}

@@

\end{exe}} For empirical evidence that at-issueness predicts projection, see AUTHORS 2018 \jd{just cite t et al}, who investigated the hypothesis for 19 projective contents associated with American English expressions.

@@

The proposal not only accounts for the projectivity of the prejacent of EAS but also for the interaction in the projection of the prejacent and the generalization. Given that both the prejacent and the generalization can be at-issue, but exactly one of them may be at-issue in any given utterance of an EAS, it falls out from \jd{I think "from" should be "of" here and in other places? not sure about my L2 intuitions} the analysis that if the prejacent projects, the generalization does not, and vice versa: if the prejacent is not at-issue, it projects, and hence the generalization must be at-issue and doesn't project.

@@

The next section provides experimental evidence in support of the question-based projection analyses of EAS. Specifically, it presents the findings of two experiments designed to investigate two predictions of the analysis: the first prediction is that the stronger the inference to the truth of the generalization from the common ground, the less likely the prejacent is taken to project. The second prediction is that the stronger the inference to the truth of the generalization from the common ground, the more likely the prejacent is taken to be not at-issue. \jd{here the predictions are formulatetd in probabilistic terms, but the way the theory is framed above is in absolute terms: if one projects, the other doesn't. is the idea to stick to the absolute view in any given instance, but allow for uncertainty (and consequently, gradedness in behavior) in identifying the discourse question? also, can we shorten and make the predictions easier to read/follow, eg like so:}

\jd{Predictions: The stronger the inference to the truth of the generalization from the common ground,
\begin{enumerate}
\item the less likely the prejacent is taken to project.
\item the more likely the prejacent is taken to be not at-issue.
\end{enumerate}}

@@

The question-based projection analysis we developed in the previous section leads us to expect that the more the prejacent or the generalization are not at-issue, the more they project. Preliminary evidence that the at-issueness of the prejacent influences its projection comes from AUTHORS 2018 \jd{cite}. One of the contents they investigated was the prejacent of EASs with {\em stupid} embedded under the polar question operator. The four evaluative adjective polar questions studied in their work are given in (\ref{stupid2}):

@@

\jd{the whole following discussion of karttunen feels weirdly disjoint from the discussion of tonhauser et al 2018 above, even though both are framed as preliminary evidence for the analysis and as motivation for the current exp. maybe move the karttunen discussion up to the exp 1 intro? } Preliminary evidence for this prediction comes from an experiment reported on in \citealt{karttunen-etal2014}. In that experiment, participants were presented with written NEASs like those in (\ref{paul}), referred to as `statements', and they were asked whether the author of the statement believes the prejacent or the negation of the prejacent; a third response option was `cannot decide'. The experiment included one triple like that in (\ref{paul}) for each of the 19 evaluative adjectives tested ({\em arrogant, brave, careless, cruel, evil,
foolish, fortunate, heroic, humble, lucky, mean, nice, polite, rude,
sensible, smart, stupid, sweet, wise}). Each triple included a NEAS referred to by \citet{karttunen-etal2014} as `consonant', which means that ``there is a predisposition to assume or grant that for NP to VP would be Adj'' (p.237). For instance, the NEAS in (\ref{paul}a) is consonant because for Paul to take the best piece is smart; this is comparable to what we have characterized as the generalization following from the common ground. Each triple also included a NEAS that \citet{karttunen-etal2014} referred to as `dissonant', which means that ``there is a predisposition to assume or grant that for NP to VP would not be Adj'' ({\em ibid}). The NEAS in (\ref{paul}c) is dissonant because for the man to take the worst piece is not smart; this is comparable to what we have characterized as the falsity of the generalization following from the common ground. Finally, the third NEAS in each triple was considered `neutral', i.e., neither consonant or dissonant, like (\ref{paul}b).

@@

A second goal of Exp.~1 was to explore the prediction of our analysis that the discourse status of the generalization is influenced not just by the truth-conditional content of the EAS but also by contextual information.\jd{is that really a direct prediction?} For this reason, Exp.~1 manipulated whether the critical information about the discourse status of the generalization came from the EAS or from context. Exp.~1 collected participants' ratings of the projection of the prejacents of NEASs. Specifically, the experiment collected gradient projection ratings, under the assumption that projection is a gradient property of utterance content, because speakers can be taken to be more or less committed to the truth of a particular content (AUTHORS 2018).\footnote{As discussed in AUTHORS: 498f., there are at least two interpretations of what it means for projectivity to be a gradient property: ``On a first interpretation, a listener's (or reader's) judgment that a content is projective to a certain extent means that the listener takes the speaker (or writer) to be committed to the content to that extent. On this interpretation, projectivity being a gradient property is a consequence of speaker commitment being a gradient property. On a second interpretation, a listener's judgment that a content is projective to a certain extent reflects the probability with which they believe the speaker to be committed to the content. On this interpretation, speaker commitment may be a binary, categorical property and projection variability arises from the listener?s uncertainty about the whether the speaker is committed." Like AUTHORS 2018, we remain agnostic about the underlying interpretation of projectivity as a gradient property, though our discussion of projection variability will be in line with the first interpretation.}

@@

To assess whether participants were attending to the task, the same 6 control stimuli were added to each list, for a total of 16 stimuli per list. As shown in (\ref{control-exp1}), the control stimuli also consisted of two-sentence discourses: the content of interest in (\ref{control-exp1}a-c) was hypothesized to be true, and the content of interest in (\ref{control-exp1}d-f) was hypothesized to be false. \jd{put these sentences in the appendix}

@@

According to the analysis of EASs developed in section \ref{s2}, the prejacent is not a lexically-specified presupposition but projects when it is not at-issue with respect to the Discourse Question of the utterance. The analysis predicts that the more the truth of the generalization follows from the common ground \jd{the more it follows or the more likely it is to follow?}, the more likely it is that the Discourse Question is not about the generalization, but about the prejacent, and, hence, the more likely the prejacent is to not project. In this section, we provided empirical evidence for two predictions of this analysis:  Exp.~1 showed that the prejacent is more likely to project when the truth of the generalization does not follow from the common ground than when the truth of the generalization does not follow from the common ground, and Exp.~2 showed that the prejacent is more likely to be at-issue when the truth of the generalization follows from the common ground than when the truth of the generalization does not follow from the common ground.

@@

\caption{Count of participants in Experiment 1 by their mean certainty ratings for the two types of NEASs\jd{relabel y-axis "Number of participants"; update caption to "histogram of participants..."}}\label{f-dialect}
\end{center}
\end{figure}

@@

\bibliographystyle{cslipubs-natbib}
\bibliography{bibliography}
